---
title: "Finding local maxima with MassSpecWavelet"
author:
- name: Sergio Oller
  affiliation:
  - &id Institute for Bioengineering of Catalonia, Barcelona, Spain
package: MassSpecWavelet
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Finding local maxima}
  %\VignetteKeywords{Peak Detection}
  %\VignetteKeywords{Wavelet}
  %\VignetteKeywords{MassSpecWavelet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

As described in the introductory vignette, one of the steps of the peak detection
process in the MassSpecWavelet package relies on the detection of local maxima
on the wavelet coefficients.

MassSpecWavelet version 1.63.3 introduced a new algorithm for detecting local
maxima that addressed some issues from the former algorithm. In this vignette:

- We show the issues of the classic `localMaximum()` detection algorithm, as well
  as how to switch between both algorithms, if you need to
- We describe the new algorithm
- We compare the performance of both


If you want to keep using the "classic" algorithm, because you care that much
about reproducibility, you can use
`options("MassSpecWavelet.localMaximum.algorithm" = "classic")`.

```{r}
library(MassSpecWavelet)
```


# Problems with the classic algorithm

The `"classic"` `localMaximum` algorithm implements the local maxima detection
using a two partially overlapping non-sliding windows: One that starts at the beginning of
the signal, and another one starting at half the window size.

The documentation reflected this behavior in the Details section:

> Instead of find the local maximum by a slide window, which slide all possible
> positions, we find local maximum by transform the vector as matrix, then get
> the the maximum of each column. This operation is performed twice with 
> vecctor shifted half of the winSize. The main purpose of this is
> to increase the efficiency of the algorithm.

While it's true that this makes the algorithm faster, it does so at the expense of
missing some peak detections. See for instance the following artificial signal:


```{r}
peak <- c(1,2,3)
x <- c(0, peak, peak, 0, 0, 0, peak, 1, peak, peak, 0)
plot(x, type = "o", main = "Five peaks are expected")
```
If we use a sliding window of four points we can detect the five peaks (if you
actually look at each maximum you will be able to draw some window that contains
the maximum and it has at least four points inside). So when we
use the new algorithm:

```{r}
options("MassSpecWavelet.localMaximum.algorithm" = "new")
local_max <- which(localMaximum(x, winSize = 4) > 0)
plot(x, type = "o", main = "With the new algorithm, 5/5 peaks are found")
points(local_max, x[local_max], col = "red", pch = 20)

```
However with the classic algorithm, there are two peaks we are missing:

```{r}
options("MassSpecWavelet.localMaximum.algorithm" = "classic")
local_max <- which(localMaximum(x, winSize = 4) > 0)
plot(x, type = "o", main = "With the classic algorithm, 3/5 peaks are found")
points(local_max, x[local_max], col = "red", pch = 20)
```

While this is less likely to happen with longer window sizes, and the default window
size in the peak detection is `2*scale+1`, this efficiency shortcut may have 
some actual impact on the peak detection of MassSpecWavelet.

I haven't measured how often this happens in real mass spectrometry samples. Feel
free to compare the results and get back to me, I guess the impact will be rather
small, otherwise this issue would have been noticed earlier.

Benchmarking the performance of the MassSpecWavelet peak detection algorithm, this
local maximum detection was also the most computationally intensive part in my
tests, so finding a more efficient (and more correct) algorithm seemed a worth
pursuing goal to improve the package.

# The new local maximum algorithm

We want an algorithm to detect local maximum using a sliding window. Since we
are putting effort in correctness, we want to define it properly.

A point in the signal will be reported as a local maximum if one of these
conditions are met:

- We can draw some window of size `winSize` that (a) contains our point (b) our
  point is not in the border of the window and (c) is the largest value of the
  window.

- If a point is part of a plateau of constant values, the center of the plateau
  will be considered as the maximum. If the plateau has an even size, it will be
  the first of the two points that could be the center of the plateau.

To give an example of these plateaus (very unlikely to happen with floating point
values) see this synthetic example and how the two algorithms behave:

```{r}
x <- c(0, 1, 2, 3, 3, 3, 3, 2, 1, 0, 3, 0, 3, 3, 3, 3, 3, 0)
x <- c(x, 0, 1, 2, 3, 3, 3, 2, 1, 0, 3, 0, 0, 0, 3, 3, 3, 0, 0)
options("MassSpecWavelet.localMaximum.algorithm" = "classic")
local_max_classic <- which(localMaximum(x, winSize = 5) > 0)
options("MassSpecWavelet.localMaximum.algorithm" = "new")
local_max_new <- which(localMaximum(x, winSize = 5) > 0)
par(mfrow = c(2, 1))
plot(x, type = "o", main = "With the classic algorithm, 2/6 peaks are found")
points(local_max_classic, x[local_max_classic], col = "red", pch = 20)
plot(x, type = "o", main = "With the new algorithm, 6/6 peaks are found")
points(local_max_new, x[local_max_new], col = "blue", pch = 20)

```

While these plateaus are corner cases in real scenarios very unlikely to happen,
it is worth having a well-defined and well-behaved implementation.

Since we want the algorithm to be as fast as possible, and we want to avoid
memory copies and R interpreter overhead, the algorithm is implemented in C.

We added 

```{r}
# Run this interactively
set.seed(5413L)
winSizes <- c(5, 31, 301)
xlengths <- c(20, 200, 2000, 20000, 200000)
out <- vector("list", length(winSizes) * length(xlengths))
i <- 0L
for (winSize in winSizes) {
    for (xlength in xlengths) {
        i <- i + 1L
        x <- round(10*runif(xlength), 1)*10
        bm <- as.data.frame(
                bench::mark(
                old = {
                    options(MassSpecWavelet.localMaximum.algorithm = "classic")
                    localMaximum(x, winSize = winSize)
                    
                },
                new = {
                    options(MassSpecWavelet.localMaximum.algorithm = "new")
                    localMaximum(x, winSize = winSize)
                    
                },
                check = FALSE,
                time_unit = "ms"
            )
        )
        out[[i]] <- data.frame(
            algorithm = as.character(bm$expression),
            median_cpu_time_ms = as.numeric(bm$median),
            xlength = xlength,
            winSize = winSize
        )
    }
}
out2 <- do.call(rbind, out)
library(ggplot2)
ggplot(out2) +
    geom_point(aes(x = xlength, y = median_cpu_time_ms, color = algorithm, shape = as.factor(winSize))) +
    scale_x_log10() +
    scale_y_log10() +
    labs(x = "signal length", y = "median cpu time (ms)", color = "Algorithm", shape = "winSize")

```



# Session Information

```{r}
sessionInfo()
```
